Primordial cardiomyocytes orchestrate myocardial morphogenesis and vascularization but are dispensable for regeneration

Version 1

Abstract

The vertebrate heart is composed of heterogeneous cardiomyocyte (CM) populations, however, the roles of distinct CM subpopulations in heart development and repair remain poorly defined. Here, using single-cell RNA sequencing analysis of adult zebrafish heart, we identified a unique CM subpopulation marked by the expression of phlda2, which is associated with anaerobic metabolism and different from mature CMs that are enriched for oxidative phosphorylation genes. We demonstrated phlda2+ cells constituted a primordial CM compartment localized between compact and trabecular muscles. Genetic ablation of phlda2+ CMs during development severely disrupted heart morphogenesis, leading to defective myocardial trabeculation and compaction, and impaired coronary vascularization. Surprisingly, despite their essential roles in development, the depletion of phlda2+ CMs didn’t impair myocardial restoration and revascularization following ventricular resection. We further found that primordial CMs failed to regenerate after either surgical amputation or genetic ablation, indicating a limited regenerative capacity. Our findings identify primordial CMs as an organizer for heart morphogenesis but not essential for regeneration, revealing a fundamental difference between developmental and regenerative programs in the vertebrate heart.

Introduction

The vertebrate myocardium is not a homogeneous tissue but consists of cardiomyocyte (CM) subpopulations that are functionally and molecularly different8,9. Recent single-cell transcriptomic studies have revealed this cellular heterogeneity in both mammalian and zebrafish hearts, showing clusters of CMs with specialized metabolic, structural, and contractile properties 10–14. Although these studies provide the evidence of cellular diversity of the myocardium, the functional significance of these CM subpopulations during heart development and disease remains a central question in cardiac biology.

Humans, like all mammals, lack natural ability to efficiently replace lost myocardium with new contractile tissue, a deficiency that results in the leading cause of morbidity and mortality15–17. However, zebrafish possess a remarkable capacity to effectively regenerate the lost CMs after injury2,6,18–21, facilitated by non-myocardial cells including epicardial cells3,5, immune cells22,23, endothelial cells24–26 and nerve cells27, making zebrafish an excellent model for exploring natural heart regeneration. Previous studies have shown that spared CMs dedifferentiate and proliferate to replace the lost myocardium2,16,28. However, it remains unclear whether CM subpopulations contribute equally to myocardial restoration. Furthermore, although the regeneration process has been generally accepted to recapitulate the developmental program29, it is unknown whether all CM subpopulations play similar roles in heart morphogenesis and regeneration. Therefore, identifying and characterizing the roles of CM subpopulations are critical for understanding the mechanisms of heart development and repair.

Adult zebrafish myocardium is composed of three main types of CMs with different spatial arrangements: the inner mass of trabecular CMs, the outer layer of compact CMs, and a single-cell-thick layer of primordial CMs that located beneath the compact layer30–32. While the trabecular and compact muscles have been extensively studied30,33–36, the function of primordial CMs remains elusive due to the lack of genetic tools for their manipulation. Primordial CMs are thought to retain immature characteristics37, but their molecular identity and functional contributions to myocardial morphogenesis and regeneration have not been fully elucidated. Here, we performed single-cell RNA sequencing (scRNA-seq) analysis of adult zebrafish heart to define CM subpopulations, and investigated the contribution of primordial CMs to myocardial morphogenesis and regeneration. Our studies reveals that primordial CMs are an essential organizer for myocardial morphogenesis and coronary vascularization, but not required for myocardial restoration and revascularization, highlighting regeneration goes beyond the reactivation of developmental programs.

Methods

Zebrafish and heart injuries

4-10-month-old outbred EK or EK/AB zebrafish were used for ventricular resection surgeries. To deplete phlda2-expressing cells, phlda2:mCherry-NTR transgenic zebrafish were used at the age of 6-8 weeks (juvenile) and 4-12 months (adult). Animal density was maintained at ∼4 fish/L in all experiments. The phlda2:mCherry-NTR animals and control wild type siblings were treated with vehicle or 10 mM metronidazole (Mtz) (MilliporeSigma) in a 1.5 L mating tank for 12 hours/day to ablate phlda2+ cells. Transgenic strains described elsewhere include: gata4:EGFP1; cmlc2:EGFP2; tcf21:mCherry3; and deltaC:EGFP4. All transgenic strains were analyzed as hemizygotes. All animal procedures were performed in accordance with Emory University guidelines.

Constructs, zebrafish transgenes

phlda2:EGFP

The phlda2 translational start codon in the BAC clone CH211-193F12 was replaced with EGFP by Red/ET recombineering technology (GeneBridges). 5′ and 3′ homologous arms flanking the EGFP cassette for recombination were a 248-base-pair (bp) fragment upstream of the start codon and a 364-bp fragment downstream, respectively. BACs for transgenesis were purified with a midi-prepare 100 kit (Qiagen) and co-injected with PI-SceI into one-cell-stage zebrafish embryos. The full name of this transgenic line is Tg (phlda2:EGFP)em21.

phlda2:mCherry-NTR

The phlda2 translational start codon in the BAC clone CH211-193F12 was replaced with mCherry-NTR by Red/ET recombineering. 5′ and 3′ homologous arms flanking the mCherry-NTR cassette for recombination were a 248-bp fragment upstream of the start codon, and a 364-bp fragment downstream, respectively. The full name of this transgenic line is Tg (phlda2:mCherry-NTR)em22.

Histological methods and quantification

Histological analyses were performed on 10 µm cryosections or whole-mount paraformaldehyde-fixed hearts. Sectioned ventricular tissues were imaged using a Zeiss LSM800 confocal scanning microscope with ZEN V3.7. F.

To quantify trabecular muscle loss, three medial longitudinal sections were selected from each heart and imaged. Single optical slices of the ventricle were acquired using a 20x objective lens (1024 × 1024 pixels) at the same ventricular position in sections from cmlc2:EGFP hearts. Trabecular muscle loss was quantified as the ratio of cmlc2:EGFP⁺ cell area to the total heart area within a region extending 100 μm from the ventricular wall. Quantification of gata4:EGFP+ expression in adult fish was performed as described5.

To quantify coronary expansion in developing hearts from deltaC:EGFP fish, whole-mounted specimens were selected, and ventricle images were captured using a 20x objective lens (1,024 x 1,024 pixels). The coronary vessel lengths were measured by ImageJ software. To quantify the vessel broken, the total number of vessel fragments were counted. To quantify coronary vessel expansion in regenerating hearts from deltaC:EGFP fish, images of ventricles were captured using a 20x objective lens (1,024 x 1,024 pixels). The GFP signals were measured in pixels by ImageJ software for signals in the injury site.

To quantify gata4:EGFP+ expression in developing hearts from gata4:EGFP fish, whole-mounted specimens were selected and ventricle images were captured using a 20x objective lens (1,024 x 1,024 pixels). The GFP signals were measured in pixels and mean intensity by ImageJ software in the injury site. To quantify gata4:EGFP+ expression in regenerating hearts from gata4:EGFP fish, images of ventricles were captured using a 20x objective lens (1,024 x 1,024 pixels). The GFP signals were measured in pixels by ImageJ software for signals in the injury site.

AFOG staining

A triple staining with Aniline blue, acid Fuchsin and Orange-G (AFOG) was performed as previously described6. Briefly, cryosections were fixed in Bouin’s solution (Sigma Aldrich) for 2 h at 60°C, and one hour at room temperature. The slides were washed for 30 min in tap water and then treated with 1% phosphomolybdic acid (Sigma Aldrich) and 0.25% phosphotungstic acid solution (Sigma Aldrich) for 5 min. They were rinsed with distilled water, and incubated for 5 min with AFOG solution containing 3 g acid Fuchsin (Sigma Aldrich), 2 g Orange G (Sigma Aldrich), 1 g Anilin Blue (Sigma Aldrich) in distilled water, with pH adjusted to 1.1. After 5-min wash with distilled water, the sections were were dehydrated in graded series of ethanol and two changes of xylene, then mounted with Permount mounting medium (Fisher Chemical). Degree of regeneration was scored in a blind-test assessment of AFOG stained sections at 30 dpa. Three sections with evident injury were selected per heart. The criteria for scoring were based on the amount of persisting collagen (blue staining) and the wound closure with a new myocardium. The hearts without persisting collagenous scar were classified as fully regenerated hearts. Hearts with scar remnants that were covered by a new myocardium was considered as incomplete regeneration. The absence of a new myocardium around the wound was considered as a blocked regeneration.

Single-cell RNA-sequencing

To prepare cmlc2+ cells for single-cell RNA-sequencing analyses, cmlc2:EGFP transgenic fish were raised to adult stages. Adult hearts were collected at four months old. The heart samples were digested with 0.26 U/mL LiberaseTM Thermolysin Medium (TM) based on a previously published protocol7. Dissociated cells were spun down and live EGFP+ cells were sorted by flow cytometry. Isolated cells were sent to Emory Integrated Genomics Core (EIGC) center for 10x single-cell RNA-sequencing. Single-cell RNA-seq libraries were prepared using the Chromium Single Cell 3’ Library & Cell Bead Kit v3.1 (Cat. No. 1000128, 1000127, 120262; 10x Genomics) according to the manufacturer’s protocol. Libraries were sequenced with an Illumina NextSeq550 using mid-output 150-cycle kits according to manufacturer specifications. The newly generated scRNA-seq data were demultiplexed, aligned, and quantified using Cell Ranger Single-Cell Software. Preliminary filtered data generated from Cell Ranger were used for downstream analysis by the Seurat R package according to standard workflow.

The scRNA-seq dataset has been submitted to Gene Expression Omnibus (GEO). To review, please use the GEO accession number GSE312374 with the token gbulwsiivdqlven.

Statistical analysis

All data are presented as mean ± standard error of the mean (SEM). All statistical analyses were performed using GraphPad Prism 7 software. The Mann-Whitney Rank-Sum test was used for assessing statistical differences between the 2 groups. Results with p values < 0.05 were considered statistically significant. Statistical details of experiments can be found in the figures and figure legends.

Results

Characterization of cardiomyocyte heterogeneity in adult zebrafish heart

To characterize the cellular heterogeneity of adult zebrafish myocardium, we performed scRNA-seq with isolated CMs. The cells activating the regulatory sequences of the CM maker cmlc2 were purified from the hearts of adult cmlc2:EGFP animals2. We applied stringent quality filtering and discarded a small number of EGFP−/cmlc2− cells, and obtained high-quality transcriptomes of 1668 CMs. Unsupervised clustering of these cells identified 3 different clusters (Fig.1A), each with a distinct gene expression patterns. Cluster 1 and 2 showed an enrichment of genes associated with mature CMs, such as ckmt2a (Fig.1B) and ckmt2b (Fig.S1A)37. However, due to their significant overlap, it remains unclear whether these two clusters represent distinct cellular identities of CMs. In contrast, Cluster 3 was marked by prominent expression of phlda2 (Fig.1C), as well as actn1 (Fig.1D) indicative of an immature cellular state37, and notch3 (Fig.S1B), reflecting elevated Notch signaling activity34. To gain insights into the biological characteristics of these cluster, we performed Gene Ontology (GO) enrichment analysis with specifically enriched transcripts in each CM cluster (Fig.1E). Clusters 1 and 2 were enriched for GO terms related to aerobic respiration and oxidative phosphorylation, suggesting the mature CMs which efficiently produce the energy to sustain the constant contraction38,39. Cluster 3 showed reduced enrichment for the metabolic pathways, suggesting an immature status. Notably, Cluster 3 was also enriched for GO terms related to muscle development and the morphogenesis of epithelial sheets. This suggests that the cells in this cluster may play a role in heart development and exhibit an epithelial-like structure with polarity.

scRNA-seq experiments reveal distinct cardiomyocyte clusters in adult zebrafish hearts.

(A) Uniform manifold approximation and projection (UMAP) clustering of cmlc2+ single-cells from adult hearts. (B-D) Feature plot of ckmt2a (B), phlda2 (C), actn1 (D), expression in cardiomyocyte clusters of adult hearts. (E) Identification of cardiomyocyte clusters based on gene ontology analysis.

The phlda2:EGFP+ cells represent primordial cardiomyocytes in adult zebrafish hearts

Primordial layer is a single-CM thickness, characteristics of the embryonic ventricle, and high notch activity30,34,37. To determine whether Cluster 2 is primordial CMs, we generated phlda2:EGFP and phlda2:mCherry-NTR BAC transgenic animals, containing sequences of 64,684 bp upstream of the phlda2 translation initiation codon and 51,710 bp downstream of the stop codon. We observed prominent fluorescence signals in the ventricular wall covering the surface of the ventricle (Fig.2A). As predicted by scRNA-seq result, phlda2:EGFP+ expression occupied a subset of cells marked by cmlc2:mCherry+ fluorescence in adult hearts (Fig.2B), indicating that phlda2+ cells are CMs. Moreover, phlda2-directed fluorescence signals did not co-localize with reporter transgenes marking coronary vessels or epicardial cells3,4 (Fig.2C and 2D). Examination of juvenile hearts revealed similar expression in the primordial CM layer (Fig. S1D), indicating that phlda2 transgenic reporters label the primordial cardiac wall throughout the development.

phlda2:EGFP+ cells specifically label primordial cardiomyocytes in adult zebrafish hearts.

(A) phlda2+ cells visualized in whole cardiac sections and whole-mount heart from adult phlda2:EGFP animals. n=8. Scale bar, 100 µm. (B) Confocal slices indicating phlda2+ cells in uninjured phlda2:EGFP;cmlc2:mCherry ventricles. White arrows represent phlda2+/cmlc2+ cells. n=8. Scale bar, 10 µm. (C) Confocal slices indicating phlda2+ cells in adult phlda2:mCherry;deltaC:EGFP ventricles. n=12. Scale bars, 20 µm. (D) Confocal slices indicating phlda2+ cells in adult phlda2:EGFP;tcf21:mCherry ventricles. n=10. Scale bars, 10 µm. All data are representative of two independent experiments.

Primordial cardiomyocytes are required for morphogenesis of myocardium and coronary vascularization

The zebrafish myocardium consists of an inner trabecular layer, a thin primordial layer, and an outer compact layer. Previous multiclonal studies revealed that primordial cardiomyocytes can give rise to trabecular CMs, which later breach the primordial layer to contribute to the compact myocardium30. It was reported that nearly 60% of CMs within both the trabecular and outer myocardium originated from primordial cells labeled at the embryonic stage and analyzed at 21 dpf40. However, these observations only demonstrate a potential lineage relationship. Whether this cell population is required for the formation or structural integrity of these myocardial layers remains unknown. A major barrier to addressing this question has been the lack of genetic tools for the specific ablation of primordial cells. To address this, we used the bacterial NTR system for inducible cell ablation41. We generated a new transgenic line and incubated phlda2:mCherry-NTR adults with the prodrug metronidazole (Mtz)41. This treatment depleted approximately 96.9% of ventricular phlda2+ cells without apparent effects on animal survival (Fig.S1E). We incubated juvenile phlda2:mCherry-NTR;cmlc2:EGFP animal and control cmlc2:EGFP siblings for 3 days with Mtz, and then assess the hearts for histological analysis at 10 and 30 days after treatment (dpt). We found primordial CM depletion led to structural abnormality in the heart. The compact muscle was disorganized when compared to controls, and trabecular muscle formation was severely impaired showing an approximately 54.7% reduction closed to the ventricular wall (Fig.3A-3B). In juvenile zebrafish ventricles, gata4:EGFP+ CMs form the initial clones of compact muscle, which then gradually expand and converge with others to eventually encapsulate the ventricle and create a contiguous wall of compact muscle35. We analyzed hearts of phlda2:mCherry-NTR;gata4:EGFP fish and identified the ablation of primordial cells showed loss of sarcomere organization compared with control siblings at 10 dpt (Fig.3C). Moreover, primordial cell-depleted hearts exhibited a 16.9% reduction in gata4:EGFP fluorescence intensity and a 48.9% reduction in the fluorescence area (Fig.3D-3E). During zebrafish heart development, the ventricle becomes vascularized after the compact myocardium has formed post-embryonically42. This process is closely associated with myocardial maturation and likely depends on proper organization of the compact layer to provide structural and molecular cues for vessel growth and integrity. To determine whether ablation of primordial CMs affects vascularization during heart development, we examined coronary vessel formation in phlda2:NTR;deltaC:EGFP fish, in which deltaC:EGFP has been utilized to visualize coronary endothelial cells4,43. We found that coronary vessels were fragmented and poorly developed in primordial cell-ablated hearts, although the total vessel length showed no reduction (Fig.3F–H). These findings indicate that primordial CMs are required for myocardial growth and vascular formation.

Primordial cardiomyocytes are required for morphogenesis of myocardium and coronary vascularization.

(A) Compact and trabecular muscle in ventricular sections from Mtz-treated cmlc2:EGFP and cmlc2:EGFP;phlda2:mCherry-NTR animals. Animals were treated with Mtz during the juvenile stage and analyzed 30 dpt. Boxed area is enlarged with high magnification. Arrows represent the compact muscle and arrowheads represent trabecular muscle. N = 6-8. Scale bar, 20 µm. (B) Quantification of cmlc2⁺ cell area relative to heart area within 100 µm of the heart wall from experiments in (A). Mann-Whitney rank-sum test. (C) Visualization of gata4+ cardiomyocytes in whole-mounted juvenile gata4:EGFP and gata4:EGFP;phlda2:mCherry-NTR hearts after Mtz treatment. Boxed area is enlarged with high magnification. Arrowheads represent EGFP signals in phlda2 cell–depleted hearts. n=6-7. Scale bars, 20 µm. (D-E) Quantification of the percentage of EGFP+ pixels (D) and EGFP intensity (E) on the ventricular surface from experiments in (D) and (E). Mann-Whitney rank-sum test. (F) Coronary vessels in whole-mounted deltaC:EGFP and deltaC:EGFP;phlda2:mCherry-NTR hearts. Animals were treated with Mtz during the juvenile stage and analyzed at 30 dpt. Arrows represent the broken vessel fragments. n=5. Scale bars, 20 µm. (G-H) Quantification of vessel length/heart area (G) and vessel fragments (H) on the ventricular surface from experiments in (F). All data are representative of three independent experiments.

Primordial cardiomyocytes are not essential for myocardial regeneration and coronary revascularization

Our data indicated primordial CMs play essential roles during heart development. Next, we examined their contribution to heart regeneration. The gata4 expression is induced in regenerating CMs and represent the proliferating CMs with reactivation of a cardiac developmental program2,5,35. We treated adult phlda2:mCherry-NTR;gata4:EGFP fish and control gata4:EGFP siblings with Mtz for 12 hours per day over three consecutive days, starting 6 days before resection of the ventricular apex, and collected hearts at 7 days post-amputation (dpa). We imaged and quantified gata4:EGFP+ signals at the injury site and didn’t detect significant difference in EGFP+ signals between phlda2+cell-depleted fish and control siblings (Fig.4A-4B). In addition, with cmlc2:EGFP animals, we found that ablation of primordial CMs did not affect the regeneration index score. Myocardial regeneration proceeded normally, with new muscle effectively replacing the injured tissue by 30 dpa (Fig.4C-4D). To further assess the requirement of primordial CMs for coronary revascularization, we depleted primordial cells in phlda2:NTR;deltaC:EGFP fish and didn’t observe found obvious difference of coronary vessel density in the regenerating area of control siblings (Fig.4E-4F). Previous studies have shown that zebrafish hearts can fully regenerate without remaining scars by 30 dpa. We also performed AFOG staining, which labels intact muscle in orange, collagen in blue, and fibrin in red6,44. Results showed minimal scar area in primordial cell-depleted hearts, with no obvious difference compared to sibling controls at 30 dpa (Fig.4G-4H). Overall, these findings demonstrate that primordial CMs are not essential for heart regeneration, despite their critical role in development.

Primordial cardiomyocytes are not essential for myocardial regeneration and coronary revascularization.

(A) Section images of Mtz-treated gata4:EGFP or gata4:EGFP;phlda2:mCherry-NTR ventricles at 7 dpa assessed for gata4:EGFP+ CMs in the injury site. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 8. Scale bars, 20 µm. (B) Quantification of EGFP+ pixels from experiments in (A). Scale bars, 20 µm. Mann-Whitney rank-sum test. (C) Section images of Mtz-treated cmlc2:EGFP or cmlc2:EGFP;phlda2:mCherry-NTR ventricles at 30 dpa assessed for muscle recovery. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 9 - 10. Scale bar, 20 µm. (D) Quantification of regeneration indices from experiments in (C). Myocardial regeneration is categorized as follows: 1 = complete regeneration of a new myocardial wall; 2 = partial regeneration; and 3 = a strong block in regeneration. (E) Section images of Mtz-treated deltaC:EGFP or deltaC:EGFP;phlda2:mCherry-NTR ventricles at 30 dpa assessed for coronary vessels in the injury site. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 5 - 6. Scale bars, 20 µm. (F) Quantification of EGFP+ pixels from experiments in (E). Scale bars, 20 µm. Mann-Whitney rank-sum test. (G) Section images of Mtz-treated wt siblings or phlda2:mCherry-NTR ventricles at 30 dpa assessed for AFOG staining collagen/fibrin deposition in the injury site. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 9. Scale bars, 20 µm. (H) Quantification of scar score from experiments in (G). The scar score is categorized as follows: 1 = hearts without persisting collagenous scar; 2 = hearts with scar remnants that were covered by a new myocardium; 3 = the absence of a new myocardium around the wound. The Chi-squared test was performed. All data are representative of three independent experiments.

Primordial cardiomyocytes exhibit limited regenerative capacity

Previous studies have indicated that in adult zebrafish heart, existing CMs can robustly regenerate within several weeks after resection surgery or genetic ablation18. We next assessed the regeneration capacity of primordial CMs. We first examined the phlda2:mCherry-NTR hearts at various time points after partial resection of the ventricular apex. The primordial layers were lost at the injury site, even 60 dpa at which time the heart regeneration was complete (Fig.5A). We next asked whether this limited regenerative capacity reflects an intrinsic limitation of the cells themselves. To test it, we treat the phlda2:mCherry-NTR animals without resection with Mtz to ablate primordial CMs. We didn’t observe the recovery of phlda2:mCherry fluorescence, indicating the failure of the primordial regeneration (Fig.5B). Since primordial CMs did not regenerate after either apical resection or direct ablation, we therefore asked whether a regenerative cue could overcome this limitation. We performed apical resection after the ablation of primordial CMs, and found that primordial CMs still failed to regenerate (Fig.5C), suggesting that their limited regenerative capacity is intrinsic rather than environment dependent. Finally, we found that phlda2:EGFP+ cells in the injury site did not colocalize with gata4:EGFP+ cells, suggesting they have limited proliferative capacity and do not contribute to the new muscle (Fig. 5D). We conclude that the primordial CMs have intrinsically limited regenerative potential, explaining its absence and functional dispensability in the regenerated heart.

Primordial cardiomyocytes have limited regenerative capacity.

(A) Section images of phlda2:mCherry-NTR ventricles without injury and during regeneration at 7, and 60 dpa. n =11-13. Dashed line indicates amputation plane. Scale bar, 20 μm. (B) Section images of phlda2:mCherry-NTR ventricles without Mtz treatment and after 7 or 30 days of Mtz treatment. n = 12-15. Scale bar, 20 μm. (C) Section images of ventricles of vehicle- or Mtz-treated phlda2a:mCherry-NTR at 30 dpa. Animals were treated with Mtz during the juvenile stage and amputated as adults. Dashed line indicates amputation plane. n = 12. Scale bars, 20 µm. (D) Section images of phlda2:mCherry-NTR;gata4:EGFP ventricles at 7 dpa assessed for phlda2+ cells and gata4:EGFP+ CMs in the injury site. Box area is shown in higher magnification. Arrows represent phlda2+ cells; arrowheads represent gata4+ cells. Dashed line indicates amputation plane. n = 12. Scale bar, 20 µm. All data are representative of two independent experiments.

Discussion

In this study, we characterized CM subpopulations in the adult zebrafish heart using scRNA-seq and identified phlda2 as a marker of primordial CMs, which exhibit transcriptional features of an immature cellular state and functions associated with morphogenesis of an epithelial sheet. Using newly generated phlda2 reporter and ablation lines, we revealed that primordial CMs reside beneath the compact layer and play essential roles in ventricular morphogenesis and coronary vascularization. However, unlike other CM subtypes, these cells display very limited regenerative capacity and are dispensable for myocardial regeneration. Our findings identified primordial CMs as a crucial organizer for heart morphogenesis but plays no roles during regeneration, revealing a difference between developmental and regenerative programs in the vertebrate heart.

During heart development, multicolor clonal analyses have revealed that the primordial CM lineage contributes to trabecular formation, and trabecular CMs later breach this layer to form the compact myocardium 30,45,46. However, direct functional evidence has been lacking due to the absence of specific genetic tools. In this study, we generated a phlda2:mCherry-NTR line to specifically deplete primordial CMs. The results show that the depletion of these cells caused a marked reduction in the outer trabecular muscle, whereas inner trabeculae remained largely intact, thereby supporting previous reports that primordial CMs, positioned outside the trabecular layer, contribute to the formation of the trabeculae30. Furthermore, we observed a disorganized compact myocardium and disrupted gata4 expression after the loss of primordial CMs, indicating a crucial role for primordial CMs in coordinating myocardial structural organization and maturation. The loss of these cells likely disrupts the structural and molecular cues required for compact muscle maturation, resulting in an immature ventricular wall.

Coronary vessel formation was also severely affected by the loss of primordial CMs, leading to fragmented and poorly developed vasculature. The coronary vasculature normally invades the compact myocardium during post-embryonic development, a process tightly coupled with myocardial maturation24,42. Thus, the vascular defects we observed could arise indirectly from the disorganization of the compact myocardium, which fails to provide a proper scaffold for vessel growth. Alternatively, primordial CMs themselves may produce paracrine factors that promote vessel growth and stability. Consistently, our scRNA-seq analysis revealed that vegfaa was highly expressed in primordial CMs (Fig. S1C). VEGF signaling plays a pivotal role in vascular development, acting as an important endothelial mitogen that stimulates endothelial cell tubulogenesis, and maintains vessel stability and integrity25,47,48. These findings suggest that primordial CMs may serve as a critical architectural and signaling hub to coordinate the morphogenesis of both myocardial and vascular structures during heart development.

In contrast to their essential role in heart development, primordial CMs appear to play no role in cardiac regeneration. Depletion of these cells did not affect myocardial or vascular regeneration, nor did it impair scar resolution. Furthermore, primordial CMs failed to regenerate after either surgical amputation or genetic ablation, even under strong regenerative conditions. Their inability to regenerate in the injured tissue, together with the lack of overlap with gata4⁺ proliferative CMs, indicates that these cells possess intrinsically limited proliferative capacity. They are different from trabecular and compact CMs, which efficiently contribute to the formation of new myocardium. This intrinsic restriction likely explains why the loss of primordial CMs does not impair the heart regeneration, suggesting that zebrafish cardiac regeneration is primarily driven by other CM subsets with greater regenerative potential.

Our study demonstrates that primordial CMs are essential for heart morphogenesis but dispensable for regeneration. These findings emphasize the functional diversity among CM subtypes and broaden our knowledge of CM heterogeneity in the zebrafish heart. Importantly, our results show that heart regeneration does not simply reactivate the developmental programs across all cardiomyocytes. Instead, regeneration depends on the CM subsets with high proliferative and regenerative capacity, rather than engaging all CMs equally. This work provides new insights into heart development and regeneration and may inspire new ideas to enhance cardiac repair in mammals.

eLife Assessment

This important work identifies phlda2 as a specific marker for primordial cardiomyocytes in the adult zebrafish heart and demonstrates their essential role in myocardial morphogenesis and coronary vascularization, but not in heart regeneration. The conclusions are well supported by single-cell transcriptomics, new genetic tools, and cell-specific ablation experiments. Overall, the evidence is solid and provides insight into the difference between developmental and regenerative cardiac programs. This work will be of interest for those studying cardiac development and regeneration.

Reviewer #1 (Public review):

Summary:

This manuscript addresses an important question in cardiac biology: whether distinct cardiomyocyte (CM) subpopulations play specialized roles during heart development and regeneration. Using single-cell RNA sequencing and newly generated genetic tools, the authors identify phlda2 as a specific marker of primordial cardiomyocytes in the adult zebrafish heart. They further show that these primordial CMs function are essential for myocardial morphogenesis and coronary vascularization but are dispensable for myocardial regeneration or revascularization after injury. These findings indicate that heart regeneration doesn't simply recapitulate developmental processes.

Strengths:

A major strength of the study is the generation of a phlda2 BAC reporter, which provides a specific and reliable marker for primordial cardiomyocytes. The lack of genetic tools has previously limited functional analysis of this CM population. By using phlda2 regulatory elements to generate reporter and NTR-based ablation lines, the authors can visualize and selectively manipulate primordial CMs in vivo. This enables a direct functional interrogation rather than relying on lineage tracing or correlative evidence. Through genetic ablation, the authors convincingly demonstrate that primordial CMs are essential for myocardial morphogenesis and coronary vascular organization during development but are not necessary for heart regeneration.

Weaknesses:

(1) The manuscript would benefit from clarifying whether the primordial cardiomyocytes ablation affects epicardial cell behaviors during heart development, given that the well-established role of the epicardium in supporting coronary vessel growth, it is possible that the vascular phenotypes observed after primordial CM ablation may be affected, at least in part, by altered epicardial cells.

(2) Because primordial cardiomyocytes form a dense, single-cell-thick layer covering the ventricular surface, it would be informative to determine whether their loss alters the spatial distribution or inward migration of coronary endothelial cells or epicardial cells.

(3) The manuscript carefully examines the relationship between primordial CMs and gata4⁺ cardiomyocytes during regeneration. However, their relationship during heart development should be more fully addressed.

(4) As loss of cardiomyocytes is known to induce gata4:GFP activation during regeneration, it would be important to determine whether ablation of primordial cardiomyocytes alone triggers gata4:GFP expression in neighboring cardiomyocytes. This analysis would further support the conclusion that primordial cardiomyocytes are not required for regenerative responses.

Reviewer #2 (Public review):

Summary:

In the manuscript "Primordial Cardiomyocytes orchestrate myocardial morphogenesis and vascularization but are dispensable for regeneration", Sun et al. identify a novel marker of primordial cardiomyocytes and use it to visualize and ablate the population during development and regeneration. The role of the primordial layer has not been investigated because the tools to manipulate this population have not existed. The manuscript is straightforward, easy to understand, and addresses an important question that has not been explored.

While the manuscript provides important insights into the role of primordial CMs, backed by a convincing methodology, the authors should clarify their requirements for heart development and maturation. Specifically, is the primordial layer required for the fish to survive? Do primordial CMs regenerate when ablated during development, and do the defects observed (in trabecular and compact CMs and coronary vessels) resolve after 10 days post-treatment when they were detected?

Strengths:

The major strengths are the identification of a marker that enables manipulation of primordial cardiomyocytes and the tools generated by the team.

Weaknesses:

The major weakness is not considering the longer-term consequences of primordial layer ablation during development, as it is unclear whether the animals succumb to the acute cardiac defects observed or fully recover.

Reviewer #3 (Public review):

Summary:

The authors performed single-cell RNA sequencing of adult zebrafish hearts and identified markers for distinct cardiomyocyte subpopulations. One marker, phlda2, marks primordial cardiomyocytes. They generated transgenic reporter lines to characterize phlda2 expression patterns and a phlda2-NTR ablation line to determine the functional requirement of primordial cardiomyocytes during heart regeneration. They found that phlda2+ primordial cardiomyocytes are essential for myocardial morphogenesis and coronary vessel development. Interestingly, when phlda2+ primordial cardiomyocytes are ablated during heart regeneration, gata4+ cortical cardiomyocytes, coronary vessel revascularization, and scar tissue formation are not affected.

Strengths:

The authors identified a new primordial cardiomyocyte marker, phlda2. They further demonstrated that primordial cardiomyocytes are important for heart morphogenesis but dispensable for heart regeneration. Their findings reveal a potential difference between heart development and regeneration programs.

Weakness:

Despite the interesting findings, the authors did not provide supplemental data for their scRNAseq to demonstrate the data quality and support their conclusions, and some results are not well described.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This manuscript addresses an important question in cardiac biology: whether distinct cardiomyocyte (CM) subpopulations play specialized roles during heart development and regeneration. Using single-cell RNA sequencing and newly generated genetic tools, the authors identify phlda2 as a specific marker of primordial cardiomyocytes in the adult zebrafish heart. They further show that these primordial CMs function are essential for myocardial morphogenesis and coronary vascularization but are dispensable for myocardial regeneration or revascularization after injury. These findings indicate that heart regeneration doesn't simply recapitulate developmental processes.

Strengths:

A major strength of the study is the generation of a phlda2 BAC reporter, which provides a specific and reliable marker for primordial cardiomyocytes. The lack of genetic tools has previously limited functional analysis of this CM population. By using phlda2 regulatory elements to generate reporter and NTR-based ablation lines, the authors can visualize and selectively manipulate primordial CMs in vivo. This enables a direct functional interrogation rather than relying on lineage tracing or correlative evidence. Through genetic ablation, the authors convincingly demonstrate that primordial CMs are essential for myocardial morphogenesis and coronary vascular organization during development but are not necessary for heart regeneration.

Weaknesses:

(1) The manuscript would benefit from clarifying whether the primordial cardiomyocytes ablation affects epicardial cell behaviors during heart development, given that the wellestablished role of the epicardium in supporting coronary vessel growth, it is possible that the vascular phenotypes observed after primordial CM ablation may be affected, at least in part, by altered epicardial cells.

We thank the reviewer for this important suggestion. To address this possibility, we examined epicardial cells in primordial CM-ablated hearts using the epicardial marker tcf21. Surprisingly, we found that epicardial cells rapidly expanded following primordial CM ablation, with an increase already detected at 5 days post-treatment. This increase persisted through at least 30 dpt, when epicardial cell abundance remained higher than that in control hearts. These findings suggest that the coronary vessel defects are unlikely to result from a reduction in epicardial cell number. In addition, we investigated whether loss of the primordial CM layer permits abnormal inward migration of epicardial cells or coronary vessels into the myocardium. However, we observed no evidence of ectopic localization of either cell population following primordial CM ablation. While we cannot exclude the possibility that altered epicardial function or signaling contributes to the vascular phenotype, our results indicate that the vascular defects are not attributable to reduced epicardial cell abundance or abnormal epicardial invasion. We have included these experiments in Page 10 and 11, and Fig. 3I-3J of revised manuscript.

“Because epicardial cells play critical roles in coronary vessel development, we next asked whether the vascular defects observed following primordial CM ablation were secondary to alterations in the epicardium. We found that tcf21+ epicardial cells revealed an increase rather than a decrease in epicardial cell abundance in primordial CM-ablated hearts (Fig. 3I3J). These findings suggest that the impaired coronary vessel organization is unlikely to result from a reduction in epicardial cell number, although we cannot exclude the possibility that altered epicardial function or signaling contributes to the vascular phenotype.”

(2) Because primordial cardiomyocytes form a dense, single-cell-thick layer covering the ventricular surface, it would be informative to determine whether their loss alters the spatial distribution or inward migration of coronary endothelial cells or epicardial cells.

We appreciate the reviewer for this insightful suggestion. Because primordial cardiomyocytes form a continuous single-cell-thick layer at the ventricular surface, we examined whether their ablation affects the spatial distribution or promotes inward migration of epicardial cells or coronary endothelial cells. Using tcf21 and deltaC reporters, we carefully analyzed the localization of these cell populations within the myocardium. We did not observe any evidence of abnormal inward migration or ectopic localization of either epicardial cells or coronary endothelial cells following primordial CM ablation (Fig. 3J–3K). These results indicate that loss of the primordial CM layer does not disrupt tissue compartmentalization or lead to inappropriate cellular invasion into the myocardial interior. We have included these experiments in Page 11, and Fig. 3J-3K of revised manuscript.

“In addition, because primordial cardiomyocytes form a continuous layer at the ventricular surface, we investigated whether their ablation permits abnormal invasion of epicardial cells or coronary vessels into the myocardium. However, we observed no evidence of ectopic localization of either cell population following primordial CM ablation (Fig. 3J-3K). Thus, loss of the primordial CM layer does not appear to disrupt tissue compartmentalization or permit abnormal cellular invasion into the myocardium.”

(3) The manuscript carefully examines the relationship between primordial CMs and gata4+ cardiomyocytes during regeneration. However, their relationship during heart development should be more fully addressed.

We thank the reviewer for this important suggestion. To further address the relationship between primordial cardiomyocytes and gata4+ cardiomyocytes during heart development, we examined their spatial and cellular relationship in juvenile zebrafish hearts. Consistent with our observations during regeneration, we did not detect any overlap between phlda2+ cardiomyocytes and gata4+ cardiomyocytes in the juvenile heart (7-8 wpf). These results indicate that primordial CMs and gata4+ proliferative CMs represent distinct cardiomyocyte populations during both heart development and regeneration. We have included these experiments in Page 13, and Fig.5E of revised manuscript.

“Consistently, no overlap between phlda2+ and gata4+ cardiomyocytes was observed in the wpf juvenile zebrafish heart (Fig. 5E).”

(4) As loss of cardiomyocytes is known to induce gata4:GFP activation during regeneration, it would be important to determine whether ablation of primordial cardiomyocytes alone triggers gata4:GFP expression in neighboring cardiomyocytes. This analysis would further support the conclusion that primordial cardiomyocytes are not required for regenerative responses.

We appreciate the reviewer for this important suggestion. To determine whether ablation of primordial cardiomyocytes alone is sufficient to activate regenerative signaling, we treated adult phlda2:mCherry-NTR;gata4:EGFP fish and gata4:EGFP siblings with Mtz for 12 hours per day over three consecutive days without ventricular resection. Under these conditions, we did not observe any induction of gata4 expression following primordial CM ablation (Fig. S5). These results indicate that loss of phlda2+ cardiomyocytes alone is not sufficient to trigger regenerative gata4 activation in the absence of injury, further supporting that primordial CMs are dispensable for activation of the regenerative response. We have included these experiments in Page 11 and Fig. S5 of revised manuscript.

“To determine whether ablation of primordial cardiomyocytes is sufficient to activate regenerative signaling, we first treated adult phlda2:mCherry-NTR;gata4:EGFP fish and control gata4:EGFP siblings with Mtz for 12 hours per day over three consecutive days without ventricular resection. We did not observe any induction of gata4:EGFP expression following primordial CM ablation (Fig. S5), indicating that loss of phlda2+ cardiomyocytes is not sufficient to trigger regenerative gata4 activation in the absence of injury.”

Reviewer #2 (Public review):

Summary:

In the manuscript "Primordial Cardiomyocytes orchestrate myocardial morphogenesis and vascularization but are dispensable for regeneration", Sun et al. identify a novel marker of primordial cardiomyocytes and use it to visualize and ablate the population during development and regeneration. The role of the primordial layer has not been investigated because the tools to manipulate this population have not existed. The manuscript is straightforward, easy to understand, and addresses an important question that has not been explored.

While the manuscript provides important insights into the role of primordial CMs, backed by a convincing methodology, the authors should clarify their requirements for heart development and maturation. Specifically, is the primordial layer required for the fish to survive?

We thank the reviewer for this important question. We found that efficient ablation of phlda2+ primordial cardiomyocytes does not affect overall survival of zebrafish under standard laboratory conditions. Although these animals exhibit clear defects in cardiac structure and coronary vascular organization, they remain viable during the experimental period, indicating that the primordial CM layer is not essential for survival. While we did not assess detailed physiological parameters such as cardiac function, swimming behavior, or long-term fitness in this study, the observed structural abnormalities suggest that subtle functional consequences may exist. These aspects will be important directions for future investigation. We have included the description on page 14, paragraph 2.

“Although primordial CM ablation does not affect survival under laboratory conditions, the observed structural defects may have functional consequences on cardiac performance and overall physiological fitness, which warrant further investigation.”

Do primordial CMs regenerate when ablated during development, and do the defects observed (in trabecular and compact CMs and coronary vessels) resolve after 10 days posttreatment when they were detected?

We appreciate the reviewer for this important question. To determine whether primordial cardiomyocytes regenerate following ablation during development, we performed Mtz-mediated ablation in juvenile zebrafish (7–8 wpf) and examined the hearts at extended time points after treatment. We found that phlda2+ cardiomyocytes did not recover even at 90 days post-treatment, indicating a persistent loss of this population following developmental-stage ablation (Fig. S6). Importantly, we further assessed whether the cardiac defects observed at earlier time points resolve over time. We found that the abnormalities in trabecular and compact myocardium, as well as coronary vessel organization, persisted at 90 days post-treatment and did not show evidence of recovery (Fig. S4C). These findings demonstrate that the observed defects are not transient developmental delays but represent long-lasting structural alterations of the heart following primordial CM ablation.

Major Comments:

(1) Figure 1: A more detailed characterization of the three CM populations would be helpful in the text as well as a new Supplemental Excel Data Sheet with the top unique genes expressed in each.

We thank the reviewer for this helpful suggestion. In the revised manuscript, we have expanded the description of the three cardiomyocyte populations in the Results section to provide a more detailed characterization. We have revised the description on page 8. In addition, we have added a new Supplemental Excel Data Sheet (Table S1) listing the top differentially expressed genes for each CM cluster.

“Notably, Cluster 2 showed additional enrichment for pathways involved in mitochondrial respiratory chain assembly, ATP synthesis, TCA cycle, and ribosome biogenesis, suggesting a relatively higher metabolic and biosynthetic activity state. In contrast, Cluster 1 was enriched for GO terms associated with mitochondrial stress responses, protein degradation, and cytoprotective pathways, indicating a stress-adapted cardiomyocyte state. Cluster 3 showed reduced enrichment of metabolic pathways, consistent with an immature metabolic profile, and was further enriched for genes involved in muscle development and epithelial morphogenesis, suggesting a role in cardiac morphogenesis and tissue organization.”

(2) Figure 3: Lower magnification views of control and MTZ-treated hearts are needed for all reporters shown (cmlc2, gata4, deltaC) at 10 days post-treatment. These "whole heart" views will enable the reader to get a gross sense of how disrupted heart development is following ablation of the primordial layer.

We appreciate the reviewer for this helpful suggestion. We have added low-magnification whole-heart images for all reported markers (cmlc2, gata4, and deltaC) to better illustrate the overall cardiac morphology following ablation of the primordial layer. We have included these data in Fig. S3 of revised manuscript.

(3) It should also be stated clearly in the Results section text what day post-fertilization Mtx treatment began. A section on Mtx treatment, including timing (what day was it applied and what day was it washed out) and dose, should be added to the methods section.

We thank the reviewer for this helpful suggestion. In response, we have now clearly stated the timing of Mtz treatment in the Results section in page 9, 10 and 11 of revised manuscript.

“To address the role of phlda2+ cells during heart development, we performed the following experiments using a standardized Mtz treatment protocol (see Methods), with identical treatment conditions applied to 7-8 wpf juvenile zebrafish.”

“To address the role of primordial cells during heart regeneration, we perform the below experiments using standardized Mtz treatment protocol (see Methods), with identical treatment conditions applied to adult zebrafish (4–6 months post-fertilization).”

In addition, we have added a Mtz treatment section in the Methods, which now includes detailed information on dosage, duration, and washout schedule to ensure full reproducibility of the experiments.

“Mtz treatment

For conditional ablation of phlda2+ cardiomyocytes, zebrafish expressing phlda2:mCherryNTR were treated with 10 mM metronidazole (Mtz) for 12 hours per day for three consecutive days. Fish were washed out and maintained in fresh system water after each daily treatment. For developmental analyses, juvenile zebrafish (7–8 weeks post-fertilization) were subjected to Mtz treatment as described above. Following completion of Mtz exposure, fish were maintained under standard conditions. Hearts were collected at 10 days post-treatment for assessment of gata4 activation, and at 30 days post-treatment for analysis of myocardial structure and coronary vessel development. For regeneration experiments, adult zebrafish (4–6 months old) were similarly treated with Mtz for three consecutive days with daily washout. Three days after the final Mtz treatment, ventricular apex resection was performed. Hearts were harvested at 7 days post-amputation (dpa) for analysis of gata4 activation and early regenerative responses, and at 30 dpa for evaluation of myocardial regeneration and coronary vessel revascularization.”

(4) What happens to the heart 2 and 6 months post-treatment? Are there long-term consequences to primordial layer ablation or do the defects seen at 10 days post-treatment eventually resolve?

We thank the reviewer for this important question. To determine whether the developmental defects observed following primordial CM ablation are transient or persist long term, we performed additional analyses at later time points after Mtz washout. First, we found that phlda2+ cardiomyocytes failed to recover following Mtz-mediated ablation. In adult phlda2:mCherry-NTR fish, phlda2+ cells remained absent 30 days after Mtz washout (Fig. 5B). Similarly, when juvenile fish (7–8 wpf) were treated with Mtz and subsequently allowed to recover, phlda2+ cells were still not restored at 90 days post-treatment (Fig. S6). Second, juvenile zebrafish (7–8 wpf) were subjected to Mtz-mediated primordial CM ablation and analyzed 90 days after Mtz washout. We found that vascular abnormalities persisted long after ablation of primordial CMs (Fig. S4C). Coronary vessels remained disorganized and fragmented, indicating that the vascular phenotype does not resolve over time. Together, these findings demonstrate that primordial CM ablation causes long-lasting defects and that the abnormalities observed are not transient developmental delays. Instead, loss of primordial CMs results in persistent cellular and vascular defects that remain evident months after Mtz treatment. We have included these experiments in Fig. S4C and Fig. S6 of revised manuscript.

“Notably, these vascular abnormalities persisted at 90 days post-treatment, indicating that the defects do not resolve during subsequent cardiac growth and maturation (Fig. S4C).”

“Similarly, when juvenile zebrafish (7–8 wpf) were subjected to Mtz-mediated ablation and analyzed 90 days after treatment, phlda2+ cells remained absent, demonstrating a persistent failure of primordial CM recovery (Fig. S6).”

(5) Also, does the primordial layer come back in these animals where the lineage is ablated during development? Or is it permanently lost as shown in Figure 5 when it is ablated during adulthood?

We appreciate the reviewer for raising this important question. To determine whether the primordial layer can be reestablished following ablation during development, we treated juvenile phlda2:mCherry-NTR fish (7–8 wpf) with Mtz and examined hearts 90 days after treatment. We found that phlda2+ cardiomyocytes remained absent at this late time point (Fig. S6), indicating that the primordial layer does not recover following developmental-stage ablation. We have included these experiments in Page 12, and Fig. S6 of revised manuscript.

“Similarly, when juvenile zebrafish (7–8 wpf) were subjected to Mtz-mediated ablation and analyzed 90 days after treatment, phlda2+ cells remained absent, demonstrating a persistent failure of primordial CM recovery (Fig. S6).”

(6) Figure 4: Need to show that phlda2 reporter fluorescence in lost/reduced following Mtz treatment during adulthood before apex amputation.

We thank the reviewer for this important suggestion. We agree that confirming efficient ablation of phlda2+ cardiomyocytes in adult fish prior to regeneration analysis is essential.

In our study, we have already demonstrated in Fig.5B that Mtz treatment in adult phlda2:mCherry-NTR fish results in efficient and sustained loss of phlda2+ cells, with no detectable recovery at 7 and 30 days post-treatment. These data confirm robust ablation of the primordial CM population following Mtz treatment in adults. Therefore, additional redundant imaging prior to apex resection was not performed in Fig 4.

(7) Figure 5: It is interesting that primordial CMs do not regenerate following apex amputation or genetic ablation. This result suggests that primordial CMs are only important during development and dispensable during adulthood? This result also makes me question whether primordial CMs are actually required for heart development, which is why it is important to address whether the fish recovers.

We appreciate the reviewer for this insightful comment. Our data indicate that primordial cardiomyocytes are essential for proper heart development, as their ablation during juvenile stages leads to significant structural and vascular abnormalities. Importantly, we further examined long-term outcomes and found that these defects do not resolve over time. Juvenile zebrafish subjected to primordial CM ablation failed to recover phlda2+ cardiomyocytes even at 90 days post-treatment, and coronary vascular abnormalities also persisted at this late stage (Fig. S4C). These findings indicate that the observed developmental defects are not transient delays but instead reflect long-lasting structural alterations of the heart. In addition, we found that adult zebrafish similarly fail to regenerate primordial cardiomyocytes following genetic ablation (Fig. 5B), further supporting the limited regenerative capacity of this population. Together, these data demonstrate that primordial cardiomyocytes are required for proper cardiac development, and their loss leads to persistent defects that are not reversed during subsequent growth or regeneration.

Minor:

Line 234: Did the authors mean to write Cluster 3 (instead of Cluster 2)?

We thank the reviewer for pointing out this error. We confirm that this was a labeling mistake, and “Cluster 3” is correct. The text has been corrected in the revised manuscript.

Line 265: There is a typo of some sort in the phrase, "54.7% reduction closed to the ventricular wall".

We thank the reviewer for pointing out this error. We have changed the description on page 10 of the revised manuscript.

“The compact myocardium was disorganized compared with controls, and trabecular muscle formation was severely impaired, with an approximately 54.7% reduction in trabecular area, predominantly observed in regions adjacent to the ventricular wall (Fig. 3A, 3B and S3A).”

Is there a corollary lineage in mammals? This should be addressed in the Introduction or Discussion.

We thank the reviewer for this insightful suggestion. At present, a direct corollary lineage to zebrafish phlda2+ primordial cardiomyocytes have not been clearly defined in mammals. However, mammalian hearts also contain heterogeneous cardiomyocyte populations with distinct developmental states and metabolic profiles, including immature or embryonic-like cardiomyocytes that persist in specific regions during development and early postnatal stages. These populations may share functional similarities with the zebrafish primordial CMs in terms of developmental organization and maturation roles. We have now discussed this point in the Discussion and emphasized that whether a comparable lineage exists in mammals remains an important open question for future studies. We have included the discussion on page 15 of the revised manuscript.

“Although a direct corollary of phlda2+ primordial cardiomyocytes has not yet been identified in mammals, mammalian hearts contain heterogeneous cardiomyocyte populations with immature states. Whether these populations represent a functional equivalent of zebrafish primordial CMs remains an open question and needs further investigation.”

Reviewer #3 (Public review):

Summary:

The authors performed single-cell RNA sequencing of adult zebrafish hearts and identified markers for distinct cardiomyocyte subpopulations. One marker, phlda2, marks primordial cardiomyocytes. They generated transgenic reporter lines to characterize phlda2 expression patterns and a phlda2-NTR ablation line to determine the functional requirement of primordial cardiomyocytes during heart regeneration. They found that phlda2+ primordial cardiomyocytes are essential for myocardial morphogenesis and coronary vessel development. Interestingly, when phlda2+ primordial cardiomyocytes are ablated during heart regeneration, gata4+ cortical cardiomyocytes, coronary vessel revascularization, and scar tissue formation are not affected.

Strengths:

The authors identified a new primordial cardiomyocyte marker, phlda2. They further demonstrated that primordial cardiomyocytes are important for heart morphogenesis but dispensable for heart regeneration. Their findings reveal a potential difference between heart development and regeneration programs.

Weakness:

Despite the interesting findings, the authors did not provide supplemental data for their scRNAseq to demonstrate the data quality and support their conclusions, and some results are not well described.

We appreciate the reviewer for this important suggestion. In the revised manuscript, we have added supplemental data to support the scRNA-seq analysis, including gene expression tables for each cardiomyocyte cluster (Table S1), and full GO-term enrichment results (Table S2). In addition, we have revised the Results section to improve the clarity and description of the scRNA-seq findings. Please see detailed responses below for point-by-point clarification.

Reviewer #3 (Recommendations for the authors):

(1) The authors did not provide enough data to demonstrate the quality of their scRNAseq. They only mentioned that they obtained "high-quality" transcriptomics. Specific parameters such as how many total reads and reads per cell should be provided.

We thank the reviewer for this important suggestion. In the revised manuscript, we have added detailed sequencing quality metrics to the Methods section in Page 6 of the revised manuscript. The dataset contains 136,174,297 total reads with an average sequencing depth of 36,168 reads per cell.

“The newly generated scRNA-seq data yielded 136,174,297 total reads with an average sequencing depth of 36,168 reads per cell.”

(2) The authors utilized cmlc2:EGFP fish to perform scRNASeq. It will be helpful to include feature plots of cmlc2 and EGFP transcripts.

We thank the reviewer for this suggestion. We have now included the description in Page 8, and feature plots of cmlc2 transcripts and EGFP reporter expression in the scRNA-seq dataset as a supplementary figure (Fig. S1A and S1B).

“The expression of cmlc2 transcripts and EGFP reporter signal in the single-cell dataset further confirmed the enrichment of cardiomyocytes (Fig. S1A and S1B)”

(3) The authors show that notch 3 is in cluster 3 of cardiomyocytes and suggest that this reflects elevated NOTCH signaling activity. The authors might consider using RNAScope to further validate that Notch 3 is expressed in cardiomyocytes. It will be also helpful to confirm phlda2 expression patterns during zebrafish heart development and regeneration by RNAScope.

We appreciate the reviewer for this helpful suggestion. To further examine the spatial expression patterns of these genes, we analyzed publicly available spatial transcriptomic data from zebrafish hearts. We found that phlda2 is enriched in the outer region of the heart during both uninjured and regenerating conditions. Similarly, notch3 and actn1 were also predominantly localized to the outer heart region in the uninjured heart. These findings are consistent with our scRNA-seq results and support the spatially restricted signature of Cluster 3 cardiomyocytes. We have included these data in Page 8, and Fig. S2A-S2C of revised manuscript.

“To further validate their spatial distribution, analysis of previously published spatial transcriptomic data revealed that phlda2, notch3, and actn1 were predominantly expressed in the outer region of the heart (Fig.S2A-S2C).”

(4) The authors did not provide any data as supplemental tables to support their analyses of scRNAseq and GO-term analysis.

We thank the reviewer for this suggestion. In the revised manuscript, we have added new supplemental tables providing full support for the scRNA-seq and GO-term analyses (Table S1 and S2), including lists of differentially expressed genes for each cardiomyocyte cluster and the corresponding GO enrichment results.

(5) The description of the phenotype in Fig. 3A and B is not clear, especially for the sentence "trabecular muscle formation was severely impaired showing an approximately 54.7% reduction close to the ventricular wall.". The authors might consider using a bracket to show the compact muscle and trabecular muscle and the distance to the ventricular wall.

We appreciate the reviewer for this suggestion. We have added brackets in Fig. 3A to label the compact and trabecular myocardium for improved clarity. Regarding “distance to the ventricular wall,” we found that this measurement varies substantially across different regions within the same heart, making a single distance-based metric unreliable. Therefore, we quantified trabecular muscle using the trabecular area fraction (trabecular area/total ventricular area) within a defined region of interest as a robust and unbiased indicator. The reported 54.7% reduction refers to this area fraction, and we have revised the text accordingly for clarity in Page 10 of the revised manuscript.

“The compact myocardium was disorganized compared with controls, and trabecular muscle formation was severely impaired, with an approximately 54.7% reduction in trabecular area, predominantly observed in regions adjacent to the ventricular wall (Fig. 3A, 3B and S3A).”

(6) It is not clear how the authors quantify the vessel "length"/ventricular area and found that there is no difference (Fig. 3G). The vessel length is significantly shorter in the images (Fig. 3F) as the author also indicated that the vessels are fragmented.

We thank the reviewer for this important comment. Coronary vessel “length” was quantified by selecting a fixed region of interest (ROI) within the ventricular area, followed by skeletonization of deltaC:EGFP+ vessels using ImageJ. The total vessel length within the ROI was measured and normalized to the ROI area to obtain vessel length density. Although the representative images (Fig. 3F) show a more fragmented vascular pattern, the total summed vessel length within the defined region was not reduced. This indicates that primordial CM ablation primarily affects vascular organization rather than overall vessel length within the ventricular area. We have updated the figure legend for clarity of the revised manuscript

“Vessel length was measured within a fixed region of interest (ROI) after skeletonization of deltaC:EGFP+ vessels in ImageJ.”

Version 2

Abstract

The vertebrate heart is composed of heterogeneous cardiomyocyte (CM) populations, however, the roles of distinct CM subpopulations in heart development and repair remain poorly defined. Here, using single-cell RNA sequencing analysis of adult zebrafish heart, we identified a unique CM subpopulation marked by the expression of phlda2, which is associated with anaerobic metabolism and different from mature CMs which are enriched for oxidative phosphorylation genes. We demonstrated that phlda2+ cells constituted a primordial CM compartment localized between compact and trabecular muscles. Genetic ablation of phlda2+ CMs during development severely disrupted heart morphogenesis, leading to defective myocardial trabeculation and compaction, and impaired coronary vascularization. Surprisingly, despite their essential roles in development, the depletion of phlda2+ CMs didn’t impair myocardial restoration and revascularization following ventricular resection. We found that this was probably due to the limited regenerative capacity of the primordial CMs themselves, as they failed to regenerate after either surgical amputation or genetic ablation. Our findings identify primordial CMs as an organizer for heart morphogenesis but not essential for regeneration, revealing a fundamental difference between developmental and regenerative programs in the vertebrate heart.

Introduction

The vertebrate myocardium is not a homogeneous tissue but consists of cardiomyocyte (CM) subpopulations that are functionally and molecularly different8,9. Recent single-cell transcriptomic studies have revealed this cellular heterogeneity in both mammalian and zebrafish hearts, showing clusters of CMs with specialized metabolic, structural, and contractile properties 10–14. Although these studies provide the evidence of cellular diversity of the myocardium, the functional significance of these CM subpopulations during heart development and disease remains a central question in cardiac biology.

Humans, like all mammals, lack natural ability to efficiently replace lost myocardium with new contractile tissue, a deficiency that results in the leading cause of morbidity and mortality15–17. However, zebrafish possess a remarkable capacity to effectively regenerate the lost CMs after injury2,6,18–21, facilitated by non-myocardial cells including epicardial cells3,5, immune cells22,23, endothelial cells24–26 and nerve cells27, making zebrafish an excellent model for exploring natural heart regeneration. Previous studies have shown that spared CMs dedifferentiate and proliferate to replace the lost myocardium2,16,28. However, it remains unclear whether CM subpopulations contribute equally to myocardial restoration. Furthermore, although the regeneration process has been generally accepted to recapitulate the developmental program29, it is unknown whether all CM subpopulations play similar roles in heart morphogenesis and regeneration. Therefore, identifying and characterizing the roles of CM subpopulations are critical for understanding the mechanisms of heart development and repair.

Adult zebrafish myocardium is composed of three main types of CMs with different spatial arrangements: the inner mass of trabecular CMs, the outer layer of compact CMs, and a single-cell-thick layer of primordial CMs that located beneath the compact layer30–32. While the trabecular and compact muscles have been extensively studied30,33–36, the function of primordial CMs remains elusive due to the lack of genetic tools for their manipulation. Primordial CMs are thought to retain immature characteristics37, but their molecular identity and functional contributions to myocardial morphogenesis and regeneration have not been fully elucidated. Here, we performed single-cell RNA sequencing (scRNA-seq) analysis of adult zebrafish heart to define CM subpopulations, and investigated the contribution of primordial CMs to myocardial morphogenesis and regeneration. Our studies reveals that primordial CMs are an essential organizer for myocardial morphogenesis and coronary vascularization, but not required for myocardial restoration and revascularization, highlighting regeneration goes beyond the reactivation of developmental programs.

Results

Characterization of cardiomyocyte heterogeneity in adult zebrafish heart

To characterize the cellular heterogeneity of adult zebrafish myocardium, we performed scRNA-seq with isolated CMs. The cells activating the regulatory sequences of the CM maker cmlc2 were purified from the hearts of adult cmlc2:EGFP animals2. We applied stringent quality filtering and discarded a small number of EGFP−/cmlc2− cells, and obtained high-quality transcriptomes of 1668 CMs. The expression of cmlc2 transcripts and EGFP reporter signal in the single-cell dataset further confirmed the enrichment of cardiomyocytes (Fig. S1A and S1B). Unsupervised clustering of these cells identified 3 different clusters (Fig.1A), each with a distinct gene expression patterns. Cluster 1 and 2 showed an enrichment of genes associated with mature CMs, such as ckmt2a (Fig.1B) and ckmt2b (Fig.S1C)37. However, due to their significant overlap, it remains unclear whether these two clusters represent distinct cellular identities of CMs. In contrast, Cluster 3 was marked by prominent expression of phlda2 (Fig.1C), as well as actn1 (Fig.1D) indicative of an immature cellular state37, and notch3 (Fig.S1D), reflecting elevated Notch signaling activity34. To further validate their spatial distribution, analysis of previously published spatial transcriptomic data14 revealed that phlda2, actn1, and notch3 were predominantly expressed in the outer region of the heart (Fig.S2A-S2C). To gain insights into the biological characteristics of these cluster, we performed Gene Ontology (GO) enrichment analysis with specifically enriched transcripts in each CM cluster (Fig.1E). Clusters 1 and 2 were enriched for GO terms related to aerobic respiration and oxidative phosphorylation, suggesting the mature CMs which efficiently produce the energy to sustain the constant contraction38,39. Notably, Cluster 2 showed additional enrichment for pathways involved in mitochondrial respiratory chain assembly, ATP synthesis, TCA cycle, and ribosome biogenesis, suggesting a relatively higher metabolic and biosynthetic activity state. In contrast, Cluster 1 was enriched for GO terms associated with mitochondrial stress responses, protein degradation, and cytoprotective pathways, indicating a stress-adapted cardiomyocyte state. Cluster 3 showed reduced enrichment of metabolic pathways, consistent with an immature metabolic profile, and was further enriched for genes involved in muscle development and epithelial morphogenesis, suggesting a role in cardiac morphogenesis and tissue organization.

scRNA-seq experiments reveal distinct cardiomyocyte clusters in adult zebrafish hearts.

(A) Uniform manifold approximation and projection (UMAP) clustering of cmlc2+ single-cells from adult hearts. (B-D) Feature plot of ckmt2a (B), phlda2 (C), actn1 (D), expression in cardiomyocyte clusters of adult hearts. (E) Identification of cardiomyocyte clusters based on gene ontology analysis.

The phlda2:EGFP+ cells represent primordial cardiomyocytes in adult zebrafish hearts

Primordial layer is a single-CM thickness, characteristics of the embryonic ventricle, and high notch activity30,34,37. To determine whether Cluster 3 is primordial CMs, we generated phlda2:EGFP and phlda2:mCherry-NTR BAC transgenic animals, containing sequences of 64,684 bp upstream of the phlda2 translation initiation codon and 51,710 bp downstream of the stop codon. We observed prominent fluorescence signals in the ventricular wall covering the surface of the ventricle (Fig.2A). As predicted by scRNA-seq result, phlda2:EGFP+ expression occupied a subset of cells marked by cmlc2:mCherry+ fluorescence in adult hearts (Fig.2B), indicating that phlda2+ cells are CMs. Moreover, phlda2-directed fluorescence signals did not co-localize with reporter transgenes marking coronary vessels or epicardial cells3,4 (Fig.2C and 2D). Examination of juvenile hearts revealed similar expression in the primordial CM layer (Fig. S4A), indicating that phlda2 transgenic reporters label the primordial cardiac wall throughout the development.

phlda2:EGFP+ cells specifically label primordial cardiomyocytes in adult zebrafish hearts.

(A) phlda2+ cells visualized in whole cardiac sections and whole-mount heart from adult phlda2:EGFP animals. n=8. Scale bar, 100 µm. (B) Confocal slices indicating phlda2+ cells in uninjured phlda2:EGFP;cmlc2:mCherry ventricles. White arrows represent phlda2+/cmlc2+ cells. n=8. Scale bar, 10 µm. (C) Confocal slices indicating phlda2+ cells in adult phlda2:mCherry;deltaC:EGFP ventricles. n=12. Scale bars, 20 µm. (D) Confocal slices indicating phlda2+ cells in adult phlda2:EGFP;tcf21:mCherry ventricles. n=10. Scale bars, 10 µm. All data are representative of two independent experiments.

Primordial cardiomyocytes are required for morphogenesis of myocardium and coronary vascularization

The zebrafish myocardium consists of an inner trabecular layer, a thin primordial layer, and an outer compact layer. Previous multiclonal studies revealed that primordial cardiomyocytes can give rise to trabecular CMs, which later breach the primordial layer to contribute to the compact myocardium30. It was reported that nearly 60% of CMs within both the trabecular and outer myocardium originated from primordial cells labeled at the embryonic stage and analyzed at 21 dpf40. However, these observations only demonstrate a potential lineage relationship. Whether this cell population is required for the formation or structural integrity of these myocardial layers remains unknown. A major barrier to addressing this question has been the lack of genetic tools for the specific ablation of primordial cells. To address this, we used the bacterial NTR system for inducible cell ablation41. We generated a new transgenic line and incubated phlda2:mCherry-NTR adults with the prodrug Mtz41. This treatment depleted approximately 96.9% of ventricular phlda2+ cells without apparent effects on animal survival (Fig.S4B). To address the role of phlda2⁺ cells during heart development, we performed the following experiments using a standardized Mtz treatment protocol (see Methods), with identical treatment conditions applied to 7-8 wpf juvenile zebrafish. We incubated juvenile phlda2:mCherry-NTR;cmlc2:EGFP animal and control cmlc2:EGFP siblings for 3 days with Mtz, and then assess the hearts for histological analysis at 10 and 30 days after treatment (dpt). We found primordial CM depletion led to structural abnormality in the heart. The compact myocardium was disorganized compared with controls, and trabecular muscle formation was severely impaired, with an approximately 54.7% reduction in trabecular area, predominantly observed in regions adjacent to the ventricular wall (Fig. 3A, 3B and S3A). In juvenile zebrafish ventricles, gata4:EGFP+ CMs form the initial clones of compact muscle, which then gradually expand and converge with others to eventually encapsulate the ventricle and create a contiguous wall of compact muscle35. We analyzed hearts of phlda2:mCherry-NTR;gata4:EGFP fish and identified the ablation of primordial cells showed loss of sarcomere organization compared with control siblings at 10 dpt (Fig.3C and S3B). Moreover, primordial cell-depleted hearts exhibited a 16.9% reduction in gata4:EGFP fluorescence intensity and a 48.9% reduction in the fluorescence area (Fig.3D-3E).

Primordial cardiomyocytes are required for morphogenesis of myocardium and coronary vascularization.

(A) Compact and trabecular muscle in ventricular sections from Mtz-treated cmlc2:EGFP and cmlc2:EGFP;phlda2:mCherry-NTR animals. Animals were treated with Mtz during the juvenile stage and analyzed 30 dpt. Boxed area is enlarged with high magnification. Arrows represent the compact muscle and arrowheads represent trabecular muscle. N = 6-8. Scale bar, 20 µm. (B) Quantification of cmlc2⁺ cell area relative to heart area within 100 µm of the heart wall from experiments in (A). Mann-Whitney rank-sum test. (C) Visualization of gata4+ cardiomyocytes in whole-mounted juvenile gata4:EGFP and gata4:EGFP;phlda2:mCherry-NTR hearts after Mtz treatment. Boxed area is enlarged with high magnification. Arrowheads represent EGFP signals in phlda2 cell–depleted hearts. n=6-7. Scale bars, 20 µm. (D-E) Quantification of the percentage of EGFP+ pixels (D) and EGFP intensity (E) on the ventricular surface from experiments in (D) and (E). Mann-Whitney rank-sum test. (F) Coronary vessels in whole-mounted deltaC:EGFP and deltaC:EGFP;phlda2:mCherry-NTR hearts. Animals were treated with Mtz during the juvenile stage and analyzed at 30 dpt. Arrows represent the broken vessel fragments. n=5. Scale bars, 20 µm. (G-H) Quantification of vessel length/heart area (G) and vessel fragments (H) on the ventricular surface from experiments in (F). Vessel length was measured within a fixed region of interest (ROI) after skeletonization of deltaC:EGFP⁺ vessels in ImageJ. (I-J) Visualization of tcf21+ epicardial cells in sectioned tcf21:nucEGFP and tcf21:nucEGFP;phlda2:mCherry-NTR hearts. Animals were treated with Mtz during the juvenile stage and analyzed at 5 and 30 dpt. n=5. Scale bars, 20 µm. (K) Coronary vessels in sectioned deltaC:EGFP and deltaC:EGFP;phlda2:mCherry-NTR hearts. Animals were treated with Mtz during the juvenile stage and analyzed at 30 dpt. n=5. Scale bars, 20 µm. All data are representative of three independent experiments.

During zebrafish heart development, the ventricle becomes vascularized after the compact myocardium has formed post-embryonically42. This process is closely associated with myocardial maturation and likely depends on proper organization of the compact layer to provide structural and molecular cues for vessel growth and integrity. To determine whether ablation of primordial CMs affects vascularization during heart development, we examined coronary vessel formation in phlda2:NTR;deltaC:EGFP fish, in which deltaC:EGFP has been utilized to visualize coronary endothelial cells4,43. We found that coronary vessels were fragmented and poorly developed in primordial cell-ablated hearts, although the total vessel length showed no reduction (Fig.3F-3H and S3C). Notably, these vascular abnormalities persisted at 90 days post-treatment, indicating that the defects do not resolve during subsequent cardiac growth and maturation (Fig. S4C). Because epicardial cells play critical roles in coronary vessel development, we next asked whether the vascular defects observed following primordial CM ablation were secondary to alterations in the epicardium. We found that tcf21+ epicardial cells revealed an increase rather than a decrease in epicardial cell abundance in primordial CM-ablated hearts (Fig. 3I-3J). These findings suggest that the impaired coronary vessel organization is unlikely to result from a reduction in epicardial cell number, although we cannot exclude the possibility that altered epicardial function or signaling contributes to the vascular phenotype. In addition, because primordial cardiomyocytes form a continuous layer at the ventricular surface, we investigated whether their ablation permits abnormal invasion of epicardial cells or coronary vessels into the myocardium. However, we observed no evidence of ectopic localization of either cell population following primordial CM ablation (Fig. 3J-3K). Thus, loss of the primordial CM layer does not appear to disrupt tissue compartmentalization or permit abnormal cellular invasion into the myocardium. Together, these findings demonstrate that primordial CMs are required for proper myocardial growth and coronary vascularization during heart development.

Primordial cardiomyocytes are not essential for myocardial regeneration and coronary revascularization

Our data indicated primordial CMs play essential roles during heart development. Next, we examined their contribution to heart regeneration. To address the role of primordial cells during heart regeneration, we perform the below experiments using standardized Mtz treatment protocol (see Methods), with identical treatment conditions applied to adult zebrafish (4–6 months post-fertilization). The gata4 expression is induced in regenerating CMs and represent the proliferating CMs with reactivation of a cardiac developmental program2,5,35. To determine whether ablation of primordial cardiomyocytes is sufficient to activate regenerative signaling, we first treated adult phlda2:mCherry-NTR;gata4:EGFP fish and control gata4:EGFP siblings with Mtz for 12 hours per day over three consecutive days without ventricular resection. We did not observe any induction of gata4:EGFP expression following primordial CM ablation (Fig. S5), indicating that loss of phlda2⁺ cardiomyocytes is not sufficient to trigger regenerative gata4 activation in the absence of injury. Then we treated adult phlda2:mCherry-NTR;gata4:EGFP fish and control gata4:EGFP siblings with Mtz for 12 hours per day over three consecutive days, starting 6 days before resection of the ventricular apex, and collected hearts at 7 dpa. We imaged and quantified gata4:EGFP+ signals at the injury site and didn’t detect significant difference in EGFP+ signals between phlda2+ cell-depleted fish and control siblings (Fig.4A-4B). In addition, with cmlc2:EGFP animals, we found that ablation of primordial CMs did not affect the regeneration index score. Myocardial regeneration proceeded normally, with new muscle effectively replacing the injured tissue by 30 dpa (Fig.4C-4D). To further assess the requirement of primordial CMs for coronary revascularization, we depleted primordial cells in phlda2:NTR;deltaC:EGFP fish and didn’t observe found obvious difference of coronary vessel density in the regenerating area of control siblings (Fig.4E-4F). Previous studies have shown that zebrafish hearts can fully regenerate without remaining scars by 30 dpa. We also performed AFOG staining, which labels intact muscle in orange, collagen in blue, and fibrin in red6,44. Results showed minimal scar area in primordial cell-depleted hearts, with no obvious difference compared to sibling controls at 30 dpa (Fig.4G-4H). Overall, these findings demonstrate that primordial CMs are not essential for heart regeneration, despite their critical role in development.

Primordial cardiomyocytes are not essential for myocardial regeneration and coronary revascularization.

(A) Section images of Mtz-treated gata4:EGFP or gata4:EGFP;phlda2:mCherry-NTR ventricles at 7 dpa assessed for gata4:EGFP+ CMs in the injury site. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 8. Scale bars, 20 µm. (B) Quantification of EGFP+ pixels from experiments in (A). Scale bars, 20 µm. Mann-Whitney rank-sum test. (C)Section images of Mtz-treated cmlc2:EGFP or cmlc2:EGFP;phlda2:mCherry-NTR ventricles at 30 dpa assessed for muscle recovery. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 9 - 10. Scale bar, 20 µm. (D) Quantification of regeneration indices from experiments in (C). Myocardial regeneration is categorized as follows: 1 = complete regeneration of a new myocardial wall; 2 = partial regeneration; and 3 = a strong block in regeneration. (E) Section images of Mtz-treated deltaC:EGFP or deltaC:EGFP;phlda2:mCherry-NTR ventricles at 30 dpa assessed for coronary vessels in the injury site. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 5 - 6. Scale bars, 20 µm. (F) Quantification of EGFP+ pixels from experiments in (E). Scale bars, 20 µm. Mann-Whitney rank-sum test. (G) Section images of Mtz-treated wt siblings or phlda2:mCherry-NTR ventricles at 30 dpa assessed for AFOG staining collagen/fibrin deposition in the injury site. Adult animals were treated with Mtz before amputation. Dashed line indicates amputation plane. N = 9. Scale bars, 20 µm. (H) Quantification of scar score from experiments in (G). The scar score is categorized as follows: 1 = hearts without persisting collagenous scar; 2 = hearts with scar remnants that were covered by a new myocardium; 3 = the absence of a new myocardium around the wound. The Chi-squared test was performed. All data are representative of three independent experiments.

Primordial cardiomyocytes exhibit limited regenerative capacity

Previous studies have indicated that in adult zebrafish heart, existing CMs can robustly regenerate within several weeks after resection surgery or genetic ablation18. We next assessed the regeneration capacity of primordial CMs. We first examined the phlda2:mCherry-NTR hearts at various time points after partial resection of the ventricular apex. The primordial layers were lost at the injury site, even 60 dpa at which time the heart regeneration was complete (Fig.5A). We next asked whether this limited regenerative capacity reflects an intrinsic limitation of the cells themselves. To test it, we treat the adult phlda2:mCherry-NTR animals without resection with Mtz to ablate primordial CMs. We didn’t observe the recovery of phlda2:mCherry fluorescence, indicating the failure of the primordial regeneration (Fig.5B). Similarly, when juvenile zebrafish (7–8 wpf) were subjected to Mtz-mediated ablation and analyzed 90 days after treatment, phlda2+ cells remained absent, demonstrating a persistent failure of primordial CM recovery (Fig. S6). Since primordial CMs did not regenerate after either apical resection or direct ablation, we therefore asked whether a regenerative cue could overcome this limitation. We performed apical resection after the ablation of primordial CMs, and found that primordial CMs still failed to regenerate (Fig.5C), suggesting that their limited regenerative capacity is intrinsic rather than environment dependent. Finally, we found that phlda2:EGFP+ cells in the injury site did not colocalize with gata4:EGFP+ cells, suggesting they have limited proliferative capacity and do not contribute to the new muscle (Fig. 5D). Consistently, no overlap between phlda2⁺ and gata4⁺ cardiomyocytes was observed in the 7-8 wpf juvenile zebrafish heart (Fig. 5E). We conclude that the primordial CMs have intrinsically limited regenerative potential, explaining its absence and functional dispensability in the regenerated heart.

Primordial cardiomyocytes have limited regenerative capacity.

(A) Section images of phlda2:mCherry-NTR ventricles without injury and during regeneration at 7, and 60 dpa. n =11-13. Dashed line indicates amputation plane. Scale bar, 20 μm. (B) Section images of phlda2:mCherry-NTR ventricles without Mtz treatment and after 7 or 30 days of Mtz treatment. n = 12-15. Scale bar, 20 μm. (C) Section images of ventricles of vehicle- or Mtz-treated phlda2a:mCherry-NTR at 30 dpa. Animals were treated with Mtz during the juvenile stage and amputated as adults. Dashed line indicates amputation plane. n = 12. Scale bars, 20 µm. (D) Section images of phlda2:mCherry-NTR;gata4:EGFP ventricles at 7 dpa assessed for phlda2+ cells and gata4:EGFP+ CMs in the injury site. Box area is shown in higher magnification. Arrows represent phlda2+ cells; arrowheads represent gata4+ cells. Dashed line indicates amputation plane. n = 12. Scale bar, 20 µm. (E) Section images of juvenile phlda2:mCherry-NTR;gata4:EGFP ventricles assessed for phlda2+ cells and gata4:EGFP+ CMs. Arrows represent phlda2+ cells; arrowheads represent gata4+ cells. n = 6. Scale bar, 10 µm. All data are representative of two independent experiments.

Discussion

In this study, we characterized CM subpopulations in the adult zebrafish heart using scRNA-seq and identified phlda2 as a marker of primordial CMs, which exhibit transcriptional features of an immature cellular state and functions associated with morphogenesis of an epithelial sheet. Using newly generated phlda2 reporter and ablation lines, we revealed that primordial CMs reside beneath the compact layer and play essential roles in ventricular morphogenesis and coronary vascularization. However, unlike other CM subtypes, these cells display very limited regenerative capacity and are dispensable for myocardial regeneration. Our findings identified primordial CMs as a crucial organizer for heart morphogenesis but plays no roles during regeneration, revealing a difference between developmental and regenerative programs in the vertebrate heart.

During heart development, multicolor clonal analyses have revealed that the primordial CM lineage contributes to trabecular formation, and trabecular CMs later breach this layer to form the compact myocardium 30,45,46. However, direct functional evidence has been lacking due to the absence of specific genetic tools. In this study, we generated a phlda2:mCherry-NTR line to specifically deplete primordial CMs. The results show that the depletion of these cells caused a marked reduction in the outer trabecular muscle, whereas inner trabeculae remained largely intact, thereby supporting previous reports that primordial CMs, positioned outside the trabecular layer, contribute to the formation of the trabeculae30. Furthermore, we observed a disorganized compact myocardium and disrupted gata4 expression after the loss of primordial CMs, indicating a crucial role for primordial CMs in coordinating myocardial structural organization and maturation. The loss of these cells likely disrupts the structural and molecular cues required for compact muscle maturation, resulting in an immature ventricular wall.

Coronary vessel formation was also severely affected by the loss of primordial CMs, leading to fragmented and poorly developed vasculature. The coronary vasculature normally invades the compact myocardium during post-embryonic development, a process tightly coupled with myocardial maturation24,42. Thus, the vascular defects we observed could arise indirectly from the disorganization of the compact myocardium, which fails to provide a proper scaffold for vessel growth. Alternatively, primordial CMs themselves may produce paracrine factors that promote vessel growth and stability. Consistently, our scRNA-seq analysis revealed that vegfaa was highly expressed in primordial CMs (Fig. S1E). VEGF signaling plays a pivotal role in vascular development, acting as an important endothelial mitogen that stimulates endothelial cell tubulogenesis, and maintains vessel stability and integrity25,47,48. These findings suggest that primordial CMs may serve as a critical architectural and signaling hub to coordinate the morphogenesis of both myocardial and vascular structures during heart development. Although primordial CM ablation does not affect survival under laboratory conditions, the observed structural defects may have functional consequences on cardiac performance and overall physiological fitness, which warrant further investigation.

In contrast to their essential role in heart development, primordial CMs appear to play no role in cardiac regeneration. Depletion of these cells did not affect myocardial or vascular regeneration, nor did it impair scar resolution. Furthermore, primordial CMs failed to regenerate after either surgical amputation or genetic ablation, even under strong regenerative conditions. Their inability to regenerate in the injured tissue, together with the lack of overlap with gata4⁺ proliferative CMs, indicates that these cells possess intrinsically limited proliferative capacity. They are different from trabecular and compact CMs, which efficiently contribute to the formation of new myocardium. This intrinsic restriction likely explains why the loss of primordial CMs does not impair the heart regeneration, suggesting that zebrafish cardiac regeneration is primarily driven by other CM subsets with greater regenerative potential.

Our study demonstrates that primordial CMs are essential for heart morphogenesis but dispensable for regeneration. Although a direct corollary of phlda2⁺ primordial cardiomyocytes has not yet been identified in mammals, mammalian hearts contain heterogeneous cardiomyocyte populations with immature states. Whether these populations represent a functional equivalent of zebrafish primordial CMs remains an open question and needs further investigation. These findings emphasize the functional diversity among CM subtypes and broaden our knowledge of CM heterogeneity in the zebrafish heart. Importantly, our results show that heart regeneration does not simply reactivate the developmental programs across all cardiomyocytes. Instead, regeneration depends on the CM subsets with high proliferative and regenerative capacity, rather than engaging all CMs equally. This work provides new insights into heart development and regeneration and may inspire new ideas to enhance cardiac repair in mammals.

Methods

Zebrafish and heart injuries

4-10-month-old outbred EK or EK/AB zebrafish were used for ventricular resection surgeries. To deplete phlda2-expressing cells, phlda2:mCherry-NTR transgenic zebrafish were used at the age of 6-8 weeks (juvenile) and 4-12 months (adult). Animal density was maintained at ∼4 fish/L in all experiments. The phlda2:mCherry-NTR animals and control wild type siblings were treated with vehicle or 10 mM metronidazole (Mtz) (MilliporeSigma) in a 1.5 L mating tank for 12 hours/day to ablate phlda2+ cells. Transgenic strains described elsewhere include: gata4:EGFP1; cmlc2:EGFP2; tcf21:mCherry3; and deltaC:EGFP4. All transgenic strains were analyzed as hemizygotes. All animal procedures were performed in accordance with Emory University guidelines.

Constructs, zebrafish transgenes

phlda2:EGFP. The phlda2 translational start codon in the BAC clone CH211-193F12 was replaced with EGFP by Red/ET recombineering technology (GeneBridges). 5′ and 3′ homologous arms flanking the EGFP cassette for recombination were a 248-base-pair (bp) fragment upstream of the start codon and a 364-bp fragment downstream, respectively. BACs for transgenesis were purified with a midi-prepare 100 kit (Qiagen) and co-injected with PI-SceI into one-cell-stage zebrafish embryos. The full name of this transgenic line is Tg (phlda2:EGFP)em21.

phlda2:mCherry-NTR. The phlda2 translational start codon in the BAC clone CH211-193F12 was replaced with mCherry-NTR by Red/ET recombineering. 5′ and 3′ homologous arms flanking the mCherry-NTR cassette for recombination were a 248-bp fragment upstream of the start codon, and a 364-bp fragment downstream, respectively. The full name of this transgenic line is Tg (phlda2:mCherry-NTR)em22.

Mtz treatment

For conditional ablation of phlda2⁺ cardiomyocytes, zebrafish expressing phlda2:mCherry-NTR were treated with 10 mM Mtz for 12 hours per day for three consecutive consecutive days. Fish were washed out and maintained in fresh system water after each daily treatment. For developmental analyses, juvenile zebrafish (7–8 weeks post-fertilization,wpf) were subjected to Mtz treatment as described above. Following completion of Mtz exposure, fish were maintained under standard conditions. Hearts were collected at 10 days post-treatment for assessment of gata4 activation, and at 30 days post-treatment for analysis of myocardial structure and coronary vessel development. For regeneration experiments, adult zebrafish (4–6 months old) were similarly treated with Mtz for three consecutive days with daily washout. Three days after the final Mtz treatment, ventricular apex resection was performed. Hearts were harvested at 7 days post-amputation (dpa) for analysis of gata4 activation and early regenerative responses, and at 30 dpa for evaluation of myocardial regeneration and coronary vessel revascularization.

Histological methods and quantification

Histological analyses were performed on 10 µm cryosections or whole-mount paraformaldehyde-fixed hearts. Sectioned ventricular tissues were imaged using a Zeiss LSM800 confocal scanning microscope with ZEN V3.7. F.

To quantify trabecular muscle loss, three medial longitudinal sections were selected from each heart and imaged. Single optical slices of the ventricle were acquired using a 20x objective lens (1024 × 1024 pixels) at the same ventricular position in sections from cmlc2:EGFP hearts. Trabecular muscle loss was quantified as the ratio of cmlc2:EGFP⁺ cell area to the total heart area within a region extending 100 μm from the ventricular wall. Quantification of gata4:EGFP+ expression in adult fish was performed as described5.

To quantify coronary expansion in developing hearts from deltaC:EGFP fish, whole-mounted specimens were selected, and ventricle images were captured using a 20x objective lens (1,024 x 1,024 pixels). The coronary vessel lengths were measured by ImageJ software. To quantify the vessel broken, the total number of vessel fragments were counted. To quantify coronary vessel expansion in regenerating hearts from deltaC:EGFP fish, images of ventricles were captured using a 20x objective lens (1,024 x 1,024 pixels). The GFP signals were measured in pixels by ImageJ software for signals in the injury site.

To quantify gata4:EGFP+ expression in developing hearts from gata4:EGFP fish, whole-mounted specimens were selected and ventricle images were captured using a 20x objective lens (1,024 x 1,024 pixels). The GFP signals were measured in pixels and mean intensity by ImageJ software in the injury site. To quantify gata4:EGFP+ expression in regenerating hearts from gata4:EGFP fish, images of ventricles were captured using a 20x objective lens (1,024 x 1,024 pixels). The GFP signals were measured in pixels by ImageJ software for signals in the injury site.

AFOG staining

A triple staining with Aniline blue, acid Fuchsin and Orange-G (AFOG) was performed as previously described6. Briefly, cryosections were fixed in Bouin’s solution (Sigma Aldrich) for 2 h at 60°C, and one hour at room temperature. The slides were washed for 30 min in tap water and then treated with 1% phosphomolybdic acid (Sigma Aldrich) and 0.25% phosphotungstic acid solution (Sigma Aldrich) for 5 min. They were rinsed with distilled water, and incubated for 5 min with AFOG solution containing 3 g acid Fuchsin (Sigma Aldrich), 2 g Orange G (Sigma Aldrich), 1 g Anilin Blue (Sigma Aldrich) in distilled water, with pH adjusted to 1.1. After 5-min wash with distilled water, the sections were were dehydrated in graded series of ethanol and two changes of xylene, then mounted with Permount mounting medium (Fisher Chemical). Degree of regeneration was scored in a blind-test assessment of AFOG stained sections at 30 dpa. Three sections with evident injury were selected per heart. The criteria for scoring were based on the amount of persisting collagen (blue staining) and the wound closure with a new myocardium. The hearts without persisting collagenous scar were classified as fully regenerated hearts. Hearts with scar remnants that were covered by a new myocardium was considered as incomplete regeneration. The absence of a new myocardium around the wound was considered as a blocked regeneration.

Single-cell RNA-sequencing

To prepare cmlc2+ cells for single-cell RNA-sequencing analyses, cmlc2:EGFP transgenic fish were raised to adult stages. Adult hearts were collected at four months old. The heart samples were digested with 0.26 U/mL LiberaseTM Thermolysin Medium (TM) based on a previously published protocol7. Dissociated cells were spun down and live EGFP+ cells were sorted by flow cytometry. Isolated cells were sent to Emory Integrated Genomics Core (EIGC) center for 10x single-cell RNA-sequencing. Single-cell RNA-seq libraries were prepared using the Chromium Single Cell 3’ Library & Cell Bead Kit v3.1 (Cat. No. 1000128, 1000127, 120262; 10x Genomics) according to the manufacturer’s protocol. Libraries were sequenced with an Illumina NextSeq550 using mid-output 150-cycle kits according to manufacturer specifications. The newly generated scRNA-seq data were demultiplexed, aligned, and quantified using Cell Ranger Single-Cell Software. The newly generated scRNA-seq data yielded 136,174,297 total reads with an average sequencing depth of 36,168 reads per cell. Preliminary filtered data generated from Cell Ranger were used for downstream analysis by the Seurat R package according to standard workflow.

The scRNA-seq dataset has been submitted to Gene Expression Omnibus (GEO). To review, please use the GEO accession number GSE312374 with the token gbulwsiivdqlven.

Statistical analysis

All data are presented as mean ± standard error of the mean (SEM). All statistical analyses were performed using GraphPad Prism 7 software. The Mann-Whitney Rank-Sum test was used for assessing statistical differences between the 2 groups. Results with p values < 0.05 were considered statistically significant. Statistical details of experiments can be found in the figures and figure legends.